--- 
title: "Working with Individual Neurons as Graph Structures" 
author: "Gregory Jefferis" 
date: "`r Sys.Date()`" 
output: rmarkdown::html_vignette 
vignette: >
  %\VignetteIndexEntry{Neurons as Graphs} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8} 
---

## Introduction

Neurons can helpfully be treated as graphs (branching trees) in which nodes connected by edges
define the morphology of the neuron. The **nat** package provides a number of
built-in functions that allow you to analyse the branching structure of graphs

http://jefferis.github.io/nat/reference/index.html#section-neurons-as-graphs-

or use aspects of the branching structure to manipulate graphs. Examples of
such functions include the `strahler_order` function which calculates the 
Strahler branch order for each node or segment in the neuron or the `spine`
function that extracts the longest path across the neuron.

More sophisticated analysis and manipulations can be carried out by converting 
`neuron` objects into `ngraph` objects.

## Neurons as Graphs
A neuron will typically be a graph in the form of a binary tree. We follow 
the convention of the Matlab [trees toolbox](http://www.treestoolbox.org/) by 
Hermann Cuntz and colleagues in treating the root (typically the soma) as the 
origin of the graph and then having directed edges leaving the root.

![neuron-as-graph](fig/neuron-graph.svg)

Nodes will be one of a:

* root
* branch point
* end point
* continutation point

Each node will have a numeric identifier or label (an arbitrary integer stored the in the 
neuron's `PointNo` field that may have come from an external source) as well
an index (an integer starting at 1 and increasing without gaps). Although these
two identifiers may often be the same, code should never rely on this being the
case. In the wild, one frequently encounters cases where e.g. the numeric labels

* have gaps because a neuron was edited
* are not in the same order as the vertices
* have some other significance e.g. are globally unique across all neurons in a
  database and therefore have large values (which may be challenging for R to
  to represent give its maxint of 2^31 - 1=`r 2^31 - 1`)

```{r, message=FALSE}
library(nat)
n=Cell07PNs[[1]]
summary(n)
```

We can extract the points as follows:

```{r}
rootpoints(n)
branchpoints(n)
endpoints(n)
```

Segments are unbranched connected sequences of nodes that terminate in a branch
point or end point.

## Built-in neuron graph functions
We will give a few examples of the use of the built-in functions that treat
neurons as graphs.

### Strahler Order
The branching structure of a neuron is commonly summarised by calculating the
[Strahler Order](https://en.wikipedia.org/wiki/Strahler_number).

```{r, fig.width=6}
n=Cell07PNs[[1]]
so=strahler_order(n)
orders=1:max(so$points)
for (i in orders) {
  plot(subset(n, so$points==i), col=i, add = i!=1, boundingbox = boundingbox(n))
}
```
Note the use of multiple calls to `plot.neuron` using the `add=FALSE` argument 
for all but the first plot. Note also the use of the `boundingbox` argument/function 
in order to ensure that the plot is set up with appropriate axes for the whole
neuron even if only part of it is plotted in the first call to `plot`. 

### Spine
You can find the longest path across a neuron using the spine function.

```{r, fig.width=6}
n=Cell07PNs[[1]]
sp=spine(n)
plot(n, col='grey')
plot(sp, add=T, col='blue')
```

`spine` has a variety of options that you can use to control the results.

### Segment graph

You can use the `segmentgraph` function to make a simplified represenation of
the branching structure of the neuron. In this object (which has class `igraph`)
each unbranched segment
in the original neuron (which might have contained many vertices) is 
collapsed to a single edge joining the branch points (which are retained).

```{r, fig.width=5, fig.height=5}
sg=segmentgraph(Cell07PNs[[1]])
plot(sg)
```

## ngraph objects

The **nat** package provides a bridge to a rich cross-platform graph library, 
**igraph**. We provide a class [`ngraph`](http://jefferis.github.io/nat/reference/ngraph.html)
that provides a thin wrapper for the `igraph` class. This looks after things 
that we might need to know about a neuron (like the 3D coordinates of each node)
while still giving access to all of the graph functions in the igraph package.

```{r}
g=as.ngraph(Cell07PNs[[1]])
class(g)
g
```
You can use functions such as 
```{r}
igraph::diameter(g)
```
to find the length of the longest path across the neuron. This is defined in 
terms of the number of intervening nodes. You can also make a graph in which 
the edge weights are the euclidean distance between the connected 3D nodes:

```{r}
gw=as.ngraph(Cell07PNs[[1]], weights=TRUE)
igraph::diameter(gw)
```
This gives you the longest path length (geodesic) across the graph in units of
µm in this case.

Note that although you can do `library(ngraph)`, it adds a lot of functions to 
the search path, some of which have name clashes, so I often just use the 
package name (`igraph::`) prepended to the function that I want to call.

### walking along ngraph objects
You can use the graph representation of neurons e.g. to find the path between
nodes.

```{r}
g=as.ngraph(Cell07PNs[[1]], weights=TRUE)
eg=endpoints(g)

p=igraph::shortest_paths(g, from=1, to=180)
p$vpath[[1]]

# fails
p2=igraph::shortest_paths(g, from=180, to=1)
p2$vpath[[1]]

# mode all will find the path irrespective of direction of links, which are
# directed from the soma
p3=igraph::shortest_paths(g, from=180, to=1, mode = 'all')
p3$vpath[[1]]
all.equal(p$vpath[[1]], rev(p3$vpath[[1]]))

# just the distances - by default uses mode=all
igraph::distances(g, v=1, to=180)
igraph::distances(g, v=180, to=1)
```

You can also ask for nodes upstream or downstream of a given starting node. For 
example the neurons in the `Cell07PNs` set have a tag called `AxonLHEP` that
defines the entry point of the axon into the lateral horn neuropil of the fly
brain. Here we defined 

```{r}
n=Cell07PNs[[1]]
g=as.ngraph(n)
# find the nodes distal to this point
# nb you must set unreachable=F if you only want to get downstream nodes
igraph::dfs(g, neimode='out', unreachable = FALSE, root=n$AxonLHEP)
# the proximal nodes back to the soma (including any branches)
igraph::dfs(g, neimode='in', unreachable = FALSE, root=n$AxonLHEP)
```

Note that `dfs` (depth first search) provides a good way to visit all the nodes
of the neuron 

Let's use this to make a function that prunes neurons downstream of this axon
entry point:

```{r, fig.width=6}
prune_from_lhep <- function(n, ...) {
  g=as.ngraph(n)
  downstream_indices=igraph::dfs(g, root = n$AxonLHEP, unreachable = FALSE)$order
  prune_vertices(n, verticestoprune = downstream_indices, invert = TRUE)
}
pruned=nlapply(Cell07PNs[1:3], prune_from_lhep)
plot(Cell07PNs[1:3], col='grey')
plot(pruned, lwd=2, add = T)
```

The pruned neurons show up in red, green, and blue in the above plot.
